The Design, Semantics, and Implementation of the Synchronous Language \CEU

### 2016-03-26 paper.tex
(+) some TODOs
...

\usepackage{listings}
%\usepackage{textcomp}
\usepackage{url}
\lstset{
%columns=fullflexible,
%basicstyle=\ttfamily,
escapeinside={||},
    %mathescape=true,
    language=C, % choose the language of the code
    basicstyle=\fontfamily{pcr}\selectfont\scriptsize\color{black},
    keywordstyle=\color{black}\bfseries, % style for keywords
    numbers=none, % where to put the line-numbers
    numberstyle=\tiny, % the size of the fonts that are used for the line-numbers
    backgroundcolor=\color{light},
    showspaces=false, % show spaces adding particular underscores
    showstringspaces=false, % underline spaces within strings
    showtabs=false, % show tabs within strings adding particular underscores
    %frame=single, % adds a frame around the code
    tabsize=2, % sets default tabsize to 2 spaces
    %rulesepcolor=\color{gray}
    captionpos=b, % sets the caption-position to bottom
    breaklines=false, % sets automatic line breaking
    %breakatwhitespace=false,
    numbersep=2em,
    % C was used in the blocksworld example to refer to block C and nowhere else
    emph={par,or,hor,do,end,loop,await,emit,input,event,call,with,%
          var,and,then,else,return,pure,deterministic,nohold,finalize,%
          class, every, FOREVER, this, spawn, in, pool, watching, until, 
          interface, each, abort, when, signal, PROC, CHAN, SIGNAL, PAR, not,
          bool, data, tag, escape, new, traverse,implementation,output},
    emphstyle={\bfseries},
    commentstyle=\color{dark}\scriptsize,
    %xleftmargin=20pt,
    %xrightmargin=20pt,
    framesep=20pt,
    %upquote=true,
    %aboveskip={1.5\baselineskip},
}

@@

Despite its success and influence, Esterel has an overly complex semantics that 
requires careful static analysis to detect and refuse programs with 
\emph{causality} and \emph{schizophrenia} problems.
Both problems have an extensive coverage and debate in the literature 
~\cite{esterel.constructive,esterel.d7,esterel.d6,esterel.d3,esterel.d5,esterel.d8,esterel.d1,esterel.schizo2}.
%
The complex semantics not only challenges the analysis and compilation of 
programs, but also results in incompatible and non-compliant implementations.
%
Furthermore, it also affects the programmer's understanding about the code, 
which, after all, has to solve the errors when facing corner cases.
%
Another drawback of the Esterel semantics consists of the loose and 
non-deterministic execution for intra-reaction statements, which prevents 
threads to share memory and interact with side-effect and stateful system calls 
appropriately.
%
For instance, although Esterel supports orthogonal abortion of 
threads~\cite{esterel.preemption}, it doesn't offer an effective mechanism to 
release resources in use.

In this work, we present \CEU, a new programming language that inherits the 
synchronous and imperative mindset of Esterel but diverges in fundamental 
semantic aspects.
%
Overall, \CEU has a simple semantics with fine-grained execution control, and a 
straightforward implementation targeting resource-constrained systems.
%
The list that follows summarizes the semantic peculiarities of our design:
%
\begin{itemize}
%
\item Stack-based execution for internal events, which provides a limited form 
of coroutines.
%
\item Static temporal analysis and deterministic execution semantics that 
allows programs to safely share memory.
%
\item Finalization mechanism for safe abortion of lines of execution holding 
external resources.
%
\item First-class synchronized timers.
\end{itemize}
%
We discuss the design of \CEU and present a formal semantics for a small 
synchronous kernel that represents a subset of the language covering these new 
functionalities.
%
We also present a lightweight implementation of \CEU with two back ends:
one aiming for resource efficiency and interoperability with \C,
and another based on a virtual machine that allows remote reprogramming.
%
Our implementations target resource-constrained devices, such as \emph{Arduino} 
and \emph{MICAz} sensor nodes based on 8-bit microcontrollers%
\footnote{
Both \emph{Arduino} and \emph{MICAz} use the 8-bit \emph{ATmega328} 
microcontroller with 32K of FLASH and 2K of SRAM.
}, showing that the peculiarities in the semantics of \CEU does not pose 
practical obstacles.%:\url{http://www.atmel.com/devices/atmega328.aspx}}

In previous work~\cite{ceu.sensys13,ceu.terra}, we employed \CEU in the context 
of wireless sensor networks, developing a number of applications, protocols, 
and drivers.
%
We evaluated the expressiveness of \CEU in comparison to event-driven code in 
\C and attested a reduction in source code size (around 25\%) with a small 
increase in memory usage (around 5--10\% for \emph{text} and 
\emph{data})~\cite{ceu.sensys13}.
%
For the \VM back end, applications have a bytecode footprint in the order of 
hundreds of bytes can be transmitted over the air in a few 
packets~\cite{ceu.terra}.

@@

In the sections that follow, we discuss the main differences between \CEU and 
Esterel:
queue-based external events and stack-based internal events 
(Section~\ref{sec.ceu.ints}),
shared-memory concurrency and determinism (Section~\ref{sec.ceu.det}),
safe abortion with finalization (Section~\ref{sec.ceu.abrt}),
and first-class timers (Section~\ref{sec.ceu.timers}).

@@

@@

\subsection{Queue-Based External Events and Stack-Based Internal Events}
\label{sec.ceu.ints}

@@

@@

%\newpage %TTT
\subsection{Shared-Memory Concurrency and Determinism}
\label{sec.ceu.det}

Embedded applications make extensive use of global memory and shared resources, 
such as through memory-mapped registers and system calls to drivers.
Hence, an important goal of \CEU is to ensure a reliable behavior for programs 
with concurrent lines of execution sharing memory and interacting with the 
environment.

Esterel is only deterministic with respect to reactive control: ``the same 
sequence of inputs always produces the same sequence of 
outputs''~\cite{esterel.primer}.
%
However, the execution order for operations within a reaction is 
non-deterministic: ``if there is no control dependency, as in \code{<<call f1() 
|| call f2()>>}, the order is unspecified and it would be an error to rely on 
it''~\cite{esterel.primer}.
%
For this reason, Esterel assumes that external calls to the host language do 
not perform side effects.
%
Also, Esterel syntactically forbids sharing memory between lines of execution 
to preserve determinism: \emph{``if a variable is written by some thread, then 
it can neither be read nor be written by concurrent 
threads''}~\cite{esterel.primer}.

Concurrency in \CEU is characterized when two or more trail segments in 
parallel execute during the same reaction.
A trail segment is a sequence of statements followed by an \code{await} (or 
termination).
%
In the example in Figure~\ref{lst.shared}.a, the two assignments to \code{x} 
(ln.  5,8) can never run concurrently, because each trail segment reacts to 
different input events (ln. 4,7), which cannot occur simultaneously (according 
to Section~\ref{sec.ceu.ints}).
%
However, the example in Figure~\ref{lst.shared}.b is non-deterministic, because 
the two assignments to \code{y} (ln. 5,8) occur in the same reaction to input 
\code{A} (ln. 4,7).

@@

\CEU performs a temporal analysis at compile time and detects concurrent 
accesses to shared variables, as follows:
if a variable is written in a trail segment, then a concurrent trail segment 
cannot read or write to that variable, nor dereference a pointer of that 
variable type.
An analogous policy is applied for pointers \emph{vs} variables and pointers 
\emph{vs} pointers, as well as for system calls with side effects (e.g., 
\code{printf}).%
\footnote{\CEU assumes that all system calls perform side effects, unless they 
are annotated as ``\texttt{@pure}''.}
%The algorithm for the analysis holds the set of all events in preceding 
%\code{await} statements for each variable access.
%Then, the sets for all accesses in parallel trails are compared to assert that 
%no events are shared among them.
%Otherwise the compiler warns about the suspicious accesses.

Regardless of the temporal analysis of \CEU, when multiple trails are active 
during the same reaction, they are scheduled in the order they appear in the 
program source code.
%
Therefore, even though the example in Figure~\ref{lst.shared}.b is suspicious, 
the assignments to \code{y} are both atomic and deterministic, i.e., after the 
reaction to \code{A} terminates, the value of \code{y} is $4$ (\code{(1+1)*2}).
%
On the one hand, enforcing an execution order for concurrent operations may 
seen arbitrary and also precludes true parallelism.
On the other hand, it provides a priority scheme for trails, and makes 
shared-memory concurrency more tractable.
%
For constrained embedded development, we believe that deterministic 
shared-memory concurrency is beneficial, given the extensive use of memory 
mapped ports for \emph{I/O} and the lack of hardware support for real 
parallelism.
Other synchronous embedded languages, such as \emph{SOL}~\cite{wsn.sol} and 
\emph{PRET-C}~\cite{rp.pretc}, made a similar design choice.
%For Esterel, however, is also used in hardware design~\cite{rp.twelve} where 
%parallelism is inherent.

Figure~\ref{lst.det} compares two syntactically equivalent code fragments in 
Esterel (a) and \CEU (b) to summarize the semantic differences regarding 
(non-)determinism.
%
Even though the program in \CEU executes deterministically, the compiler still 
issues a warning, because an apparently innocuous reordering of trails modifies 
the semantics of the program.
%
Note that in Esterel multiple external events can coexist in the same reaction, 
which disallows a similar temporal analysis.

\begin{figure}[b]
\begin{minipage}[t]{0.48\linewidth}
\begin{lstlisting}[mathescape=true]
input A;
[
    await A;
    call f1();
$\|$
    await A;
    call f2();
];
\end{lstlisting}
\centering\small{(a) Esterel}
\end{minipage}
%
\begin{minipage}[t]{0.48\linewidth}
\begin{lstlisting}
input void A;
par/and do
    await A;
    _f1();
with
    await A;
    _f2();
end
\end{lstlisting}
\centering\small{(b) \CEU}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ In Esterel, the execution order between \code{f1} and \code{f2} is 
unspecified, whereas in \CEU, \code{\_f1} executes before \code{\_f2} due to 
deterministic scheduling based on lexical order.
\label{lst.det}
}
\end{figure}

\subsection{Safe Abortion with Finalization}
\label{sec.ceu.abrt}

The introductory example in Figure~\ref{lst.abro} illustrates how synchronous 
languages can abort awaiting lines of execution without tweaking them with 
synchronization primitives.
In contrast, traditional (asynchronous) multi-threaded languages cannot express 
thread termination safely~\cite{esterel.preemption,sync_async.threadsstop}.
Still, handling abortion when dealing with external resources is challenging 
because they are not subject to the same synchronous execution discipline.

\begin{figure}
\begin{minipage}[t]{0.55\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
input void STOP, RETRANSMIT, SEND_ACK;
output _pkt_t* SEND_ENQUEUE, SEND_CANCEL;
par/or do
    await STOP;
with
    loop do
        par/or do
            await RETRANSMIT;
        with
            par/and do
                await 1min;
            with
                var _pkt_t buffer;
                <fill-buffer-info>
                emit SEND_ENQUEUE => &buffer;
                await SEND_ACK;
            end
        end
    end
end
\end{lstlisting}
\centering\small{(a)}
\end{minipage}
%
\begin{minipage}[t]{0.45\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em,firstnumber=12]
<...>
    var _pkt_t buffer;
    <fill-buffer-info>
    finalize
        emit SEND_ENQUEUE => &buffer;
    with
        emit SEND_CANCEL => &buffer;
    end
    await SEND_ACK;
<...>

.
\end{lstlisting}
\centering\small{(b)}
\end{minipage}
%\rule{8.5cm}{0.37pt}
\caption{
(a) The unsafe network protocol does not compile.
(b) The finalization clause extends the protocol to handle abortion properly.
%\caption{ Parallel compositions can describe complex state machines.
\label{lst.fin}
}
\end{figure}

>>>>

\usepackage{listings}
%\usepackage{textcomp}
\usepackage{url}
\lstset{
%columns=fullflexible,
%basicstyle=\ttfamily,
escapeinside={||},
    %mathescape=true,
    language=C, % choose the language of the code
    basicstyle=\fontfamily{pcr}\selectfont\scriptsize\color{black},
    keywordstyle=\color{black}\bfseries, % style for keywords
    numbers=none, % where to put the line-numbers
    numberstyle=\tiny, % the size of the fonts that are used for the line-numbers
    backgroundcolor=\color{light},
    showspaces=false, % show spaces adding particular underscores
    showstringspaces=false, % underline spaces within strings
    showtabs=false, % show tabs within strings adding particular underscores
    %frame=single, % adds a frame around the code
    tabsize=2, % sets default tabsize to 2 spaces
    %rulesepcolor=\color{gray}
    captionpos=b, % sets the caption-position to bottom
    breaklines=false, % sets automatic line breaking
    %breakatwhitespace=false,
    numbersep=2em,
    % C was used in the blocksworld example to refer to block C and nowhere else
    emph={par,or,hor,do,end,loop,await,emit,input,event,call,with,%
          var,and,then,else,return,pure,deterministic,nohold,finalize,%
          class, every, FOREVER, this, spawn, in, pool, watching, until, 
          interface, each, abort, when, signal, PROC, CHAN, SIGNAL, PAR, not,
          bool, data, tag, escape, new, traverse,implementation,output,
          native,@const,@pure,@safe,define},
    emphstyle={\bfseries},
    commentstyle=\color{dark}\scriptsize,
    %xleftmargin=20pt,
    %xrightmargin=20pt,
    framesep=20pt,
    %upquote=true,
    %aboveskip={1.5\baselineskip},
}

@@

% TODO: SCCharts?, SynchCharts?

Despite its success and influence, Esterel has an overly complex semantics that 
requires careful static analysis to detect and refuse programs with 
\emph{causality} and \emph{schizophrenia} problems.
Both problems have an extensive coverage and debate in the literature 
~\cite{esterel.constructive,esterel.d7,esterel.d6,esterel.d3,esterel.d5,esterel.d8,esterel.d1,esterel.schizo2}.
%
The complex semantics not only challenges the analysis and compilation of 
programs, but also results in incompatible and non-compliant implementations.
%
Above all, it also affects the programmer's understanding about the code, 
which, ultimately, has to solve the errors when facing corner cases.
%
Another drawback of the Esterel semantics consists of the loose and 
non-deterministic execution for intra-reaction statements, which prevents 
threads to share memory and interact with stateful system calls safely.
%
%For instance, although Esterel supports orthogonal abortion of 
%threads~\cite{esterel.preemption}, it doesn't offer an effective mechanism to 
%release resources in use.

In this work, we present \CEU, a new programming language that inherits the 
synchronous and imperative mindset of Esterel but diverges in fundamental 
semantic aspects.
%
Overall, \CEU has a simple semantics with fine-grained execution control, and a 
straightforward implementation targeting resource-constrained systems.
%
The list that follows summarizes the semantic peculiarities of our design:
%
\begin{itemize}
%
\item Unique and queue-based external events, which define the notion of time 
in \CEU (Section~\ref{sec.ceu.exts}).
%
\item Stack-based internal events for intra-reaction communication, which also
provides a limited form of coroutines (Section~\ref{sec.ceu.ints}).
%
\item Static temporal analysis to detect suspicious concurrent statements 
(Section~\ref{sec.ceu.analysis}).
%
\item Safe integration with \C via function annotations and enforced 
finalization for external resources (Section~\ref{sec.ceu.c}).
%
\item First-class synchronized timers with a dedicated syntax 
(Section~\ref{sec.ceu.timers}).
\end{itemize}
%
We discuss the design of \CEU and present a formal semantics for a small 
synchronous kernel that represents a subset of the language covering these new 
functionalities.
% TODO: why? stack, fin, principalmente
%
We also present a lightweight implementation of \CEU with two back ends:
one aiming for resource efficiency and interoperability with \C,
and another based on a virtual machine that allows remote reprogramming.
%
Our implementations target resource-constrained devices, such as \emph{Arduino} 
and \emph{MICAz} sensor nodes based on 8-bit microcontrollers%
\footnote{
Both \emph{Arduino} and \emph{MICAz} use the 8-bit \emph{ATmega328} 
microcontroller with 32K of FLASH and 2K of SRAM.
}, showing that the peculiarities in the semantics of \CEU do not pose 
practical obstacles.%:\url{http://www.atmel.com/devices/atmega328.aspx}}

In previous work~\cite{ceu.sensys13,ceu.terra}, we employed \CEU in the context 
of wireless sensor networks, developing a number of applications, protocols, 
and device drivers.
%
We evaluated the expressiveness of \CEU in comparison to event-driven code in 
\C and attested a reduction in source code size (around 25\%) with a small 
increase in memory usage (around 5--10\% for \emph{text} and 
\emph{data})~\cite{ceu.sensys13}.
%
For the \VM back end, applications have a bytecode footprint in the order of 
hundreds of bytes which can be transmitted over the air in a few 
packets~\cite{ceu.terra}.

@@

In the sections that follow, we discuss the main differences between \CEU and 
Esterel:
unique and queue-based external events (Section~\ref{sec.ceu.exts}),
stack-based internal events (Section~\ref{sec.ceu.ints}),
static temporal analysis for concurrent statements (Section~\ref{sec.ceu.analysis}),
safe integration with \C (Section~\ref{sec.ceu.c}), and
first-class synchronized timers (Section~\ref{sec.ceu.timers}).

@@

% TODO: input/output declarations

@@

\subsection{ Unique and Queue-Based External Events }
\label{sec.ceu.exts}

Esterel defines a logical unit of time as a discrete sequence of instants or 
``ticks''.
At each tick, depending on external stimuli the environment provides, an 
arbitrary number of input events can be active simultaneously (including zero).
The set of input events awakes the program, which must react within the current 
tick, in bounded time.

In \CEU, each unique external stimulus itself defines a unit of time.
The execution model for \CEU programs is as follows~\cite{ceu.sensys13}:
%
\begin{enumerate}
\item The program initiates the ``boot reaction'' in a single trail (but 
      parallel constructs may create new trails).
\item Active trails execute until they await or terminate.
      This step is named a \emph{reaction chain}, and always runs in bounded 
      time.
\item The program goes idle and the environment takes control.
\item On the occurrence of a new external input event, the environment awakes 
      \emph{all} trails awaiting that event.
      It then goes to step 2.
\end{enumerate}
%
A program must react completely to an occurring event completely before 
handling the next.
%
Based on the synchronous hypothesis, a program takes no time on step 2 and is 
always idle on step 3.
In practice, if a new external input event occurs while a reaction chain is 
running (step 2), it is enqueued to run in the next reaction.

\begin{figure}[!htb]
\centering
\includegraphics[width=\columnwidth]{tick}
\caption{The discrete notions of time in Esterel and \CEU.
\label{fig.ticks}
}
\end{figure}

Figure~\ref{fig.ticks} compares the discrete notions of time in Esterel and 
\CEU.
The box \code{Real World} shows event occurrences over a continuous timeline 
divided in units of 10 milliseconds.
Since Esterel and \CEU define discrete logical units of time, we describe how 
the events (which are the same in all boxes) fit into their timelines.

\code{Box-1} uses fixed-length ticks in Esterel, within which reactions to 
occurring inputs have to fit~\cite{esterel.fixed} .
%
Here, we assume a \code{R(boot)} reaction which happens before any input at 
\code{tick-0}.
%
The input \code{A} ``physically'' occurs during the boot reaction but, because 
time is logical and discrete, the event is delayed to the next tick.
%
Note that the reaction \code{R(A)} takes more time than \code{tick-1}, causing 
a \emph{timing violation} in the program execution (after the end of 
\code{tick-1} and before the end of \code{R(A)}).
%
The events \code{B} and \code{C} occur during \code{tick-1} and are delayed to 
happen \emph{simultaneously} at \code{tick-2} with \code{R(B+C)}.
%
Since no new events occurred during \code{tick-2}, the CPU stays idle during 
the whole \code{tick-3}.
%
Finally, one instance of event \code{D} and two instances of event \code{E} 
occur ``simultaneously'' during the idle \code{tick-3}.
However, because the program cannot detect both occurrences of \code{E}, only 
one is considered in \code{R(D+E)}.

\code{Box-2} is an alternative that allows variable-length ticks in Esterel 
that adjust to the time the corresponding reaction takes to 
complete~\cite{esterel.variable}.
This approach avoids the time violation for \code{R(A)} and also yields
smaller idle periods.
For instance, the occurrence of \code{D} interrupts the idle \code{tick-3} to 
start reaction \code{R(D)} on \code{tick-4}.
Similarly to the fixed-tick approach, only one of the two simultaneous 
occurrences of \code{E} is considered on \code{tick-5}.

\code{Box-3} illustrates the unique and queue-based policy for input events in 
\CEU.
We also assume a \code{R(boot)} reaction before any input.
%
Because the event \code{A} is unique during \code{tick-0}, the behavior in \CEU 
is similar to \code{Box-2} for the first two units of time.
%
However, the events \code{B} and \code{C} are not simultaneous in \CEU and are 
handled in subsequent reactions \code{R(B)} and \code{R(C)}.
%
We assume that the CPU time for \code{R(B+C)} in Esterel is roughly the same as 
\code{R(B)+R(C)} in \CEU.
This way, the first idle period in \code{Box-2} and \code{Box-3} coincide.
%
Finally, \CEU recognizes and reacts to the two instances of \code{E} 
independently, which are handled in sequence.

We decided for the unique and queue-based semantics for the reasons that 
follow:
%
\begin{itemize}
\item \textbf{A ``tick'' is too abstract and imprecise:}
Outside the domain of hardware specification, a tick has no natural counterpart 
in the real world.
Also, since no time regularity is required for ticks~\cite{esterel.multiclock}, 
the two approaches for Esterel in Figure~\ref{fig.ticks} lead to different 
behaviors for the same sequence of inputs.
%
\item \textbf{Events are never simultaneous:}
From a rigorous point of view, event occurrences are infinitesimal, having zero 
probability of being simultaneous.
This way, we believe that the notion of simultaneity should not be imposed by 
the language, but explicitly defined for each use case.
In the end of the section, we present a specification to detect 
``simultaneous'' button clicks.
%
Note that in the case of Esterel, simultaneity depends on the imprecise 
definition of ticks.
For instance, in Figure~\ref{fig.ticks}, the events \code{B} and \code{C} are
simultaneous, even though \code{A} and \code{B} actually happen much closer to 
one another.
\item \textbf{Unique input events imply mutual exclusion:}
Because they are atomic, reactions to different events never overlap.
The possibility to reason about each input individually is a prerequisite for 
the temporal analysis to be discussed in Section~\ref{sec.ceu.analysis}.
\end{itemize}

\begin{figure}
\begin{minipage}[t]{0.57\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em,mathescape=true]
input void LEFT_CLICK;
input void RIGHT_CLICK;
event void middle_click;
loop do
    par/or do
        AWAIT_AND(LEFT_CLICK, RIGHT_CLICK);
        emit middle_click;
    with
        AWAIT_OR(LEFT_CLICK, RIGHT_CLICK);
        await 200 ms;
    end
end
\end{lstlisting}
\end{minipage}
%
\begin{minipage}[t]{0.43\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em,firstnumber=13]
#define AWAIT_AND(e1, e2)   \
    par/and do              \
        await e1;           \
    with                    \
        await e2;           \
    end
#define AWAIT_OR(e1, e2)    \
    par/or do               \
        await e1;           \
    with                    \
        await e2;           \
    end
\end{lstlisting}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ Application defines that a \code{middle\_click} event occurs whenever 
both \code{LEFT\_CLICK} and \code{RIGHT\_CLICK} occur within 200 milliseconds.
\label{lst.simultaneous}
}
\end{figure}

Figure~\ref{lst.simultaneous} emulates a \code{milddle\_click} event (ln. 3) in 
terms of \code{LEFT\_CLICK} and \code{RIGHT\_CLICK} (ln. 1--2).
If both events occur, we emit the internal event \code{middle\_click} (ln.  
6--7).
However, if one of them occurs and the \emph{200ms} timer expires (ln. 9--10), 
we abort the whole behavior with the \code{par/or} (ln. 5--11) and try again 
with the enclosing \code{loop} (ln. 4).
The macros \code{AWAIT\_AND} (ln. 13--18) and \code{AWAIT\_OR} (ln. 19--24) are 
simple expansions to a \code{par/and} and \code{par/or} for better readability.
%
Here, ``simultaneous'' means ``within 200 milliseconds'', which is a huge 
amount of time for a language-defined tick, and which would break the 
synchronous hypothesis.
%
A similar implementation for Esterel would not rely on the tick notion of 
simultaneity either.
%
We discuss internal events in Section~\ref{sec.ceu.ints} and timers in 
Section~\ref{sec.ceu.timers}.

The synchronous hypothesis for \CEU holds if the reactions run faster than the 
rate of incoming input events.
Otherwise, the system will accumulate delays between the real occurrence and 
actual reaction to events.
%
This is also the case for the variable-length-tick approach of Esterel, since 
the more inputs to handle, the longer the reaction takes, and the more inputs 
will accumulate for subsequent ticks.
%
For the fixed-length-tick approach of Esterel, a breach in the synchronous 
hypothesis causes timing violations, which can be avoided with \emph{worst case 
reaction time} analysis to infer an appropriate value for the tick 
length~\cite{esterel.fixed}.

\begin{comment}
> Multiclock Esterel
The Esterel synchronous reactive language [5,4,6], which we call Classic Esterel
in this paper, is based on a single-clock instantaneous interaction principle. The
behavior of a program is defined by a sequence of reactions to input sequences.
The execution environment decides when the program is provided an input, and
a reaction is viewed as the simultaneous production of an output response to
an input event. We call ticks the instants in which reactions occur, and master
clock the sequence of these instants (it is a logical clock and no time regularity
is required).

An Esterel Processor with Full Preemption Support and its Worst Case Reaction 
Time Analysis
3. ESTEREL AND THE KIEL ESTEREL PROCESSOR
The execution of an Esterel program is divided into (logical) instants, or 
(logical) ticks, which are conceptually executed infinitely fast.
An Esterel program interacts with its environment through signals, which are 
either present throughout a logical instant or absent.
Input signals are sampled at each tick, and each tick may generate output 
signals.
It would be possible to have the system just compute one logical instant after 
the other, to start with the next reaction as soon as the previous one has 
finished; however, typically the ticks are executed at some fixed frequency, 
resulting in an interval T between each tick, where T is determined by the 
real-time requirements of the system.
The conceptually infinitely fast computation of the reaction within a logical 
tick, plus in the case of Esterel the unique signal presence/absence status 
throughout a logical tick, together constitute the synchrony hypothesis. In 
practice, the results of a logical instant of course cannot be computed 
infinitely fast; however, to maintain the abstraction of the synchrony 
hypothesis, a logical instant must be computed within some desired reaction 
time T.

4. WORST CASE REACTION TIME ANALYSIS
TODO: requires this, what about this in the context of MThreads??
Towards Direct Execution of Esterel Programs on Reactive Processors
3.1 Preserving the Synchrony Hypothesis
The synchronous model implies that reactions to input signals are instantaneous 
and occur at discrete logical instants called ticks.
As a result, a sequence of actions may need to be performed within the duration 
of a single tick.
In order to model the logical tick of Esterel in RePIC, a tick with variable 
length in terms of absolute time is implemented.
The duration of one tick is determined by the time required to execute all the 
instantaneous instructions that are placed between any two consecutive await 
(i.e. TAWAIT, SAWAIT or CAWAIT) or halt instructions which are the tick 
delimiting instructions.
(TODO: tem figuras)
\end{comment}

\subsection{ Stack-Based Internal Events }
\label{sec.ceu.ints}

@@

% TODO: .. will probably expect that duplicating the  emit inc => &v; line 
% would result in v==3.

@@

%\newpage %TTT

\subsection{ Static Temporal Analysis for Concurrent Statements }
\label{sec.ceu.analysis}

Embedded applications make extensive use of global memory and shared resources, 
such as through memory-mapped registers and system calls to device drivers.
Hence, an important goal of \CEU is to ensure a reliable behavior for programs 
with concurrent lines of execution sharing memory and interacting with the 
environment.

Esterel is only deterministic with respect to reactive control: ``the same 
sequence of inputs always produces the same sequence of 
outputs''~\cite{esterel.primer}.
%
However, the execution order for operations within a reaction is 
non-deterministic: ``if there is no control dependency, as in \code{<<call f1() 
|| call f2()>>}, the order is unspecified and it would be an error to rely on 
it''~\cite{esterel.primer}.
%
A number of Esterel-based synchronous languages, such as 
\emph{SOL}~\cite{wsn.sol},
\emph{SC}~\cite{rp.sychc}, and
\emph{PRET-C}~\cite{rp.pretc},
enforce intra-reaction determinism with an arbitrary execution order for 
statements in multiple lines of execution.
%, e.g., based on lexical order or manual priority assignment.
%
\CEU also takes the deterministic approach and when multiple trails are active 
during the same reaction, they are scheduled in the order they appear in the 
program source code.

@@

Even so, we consider that enforcing an arbitrary execution order can be 
misleading in some cases.
%
For instance, consider the two examples in Figure~\ref{lst.shared}, both of 
which define a shared variable (ln. 2), and assign to it parallel trails (ln.  
5, 8).
%
In the example \code{(a)}, the two assignments to \code{x} execute from 
reactions to different events \code{A} and \code{B} which, by definition, 
cannot occur simultaneously (Section~\ref{sec.ceu.exts}).
Hence, for the sequence \code{A->B}, \code{x} becomes \code{(1+1)*2->4}, while 
for \code{B->A}, \code{x} becomes \code{(1*2)+1->3}.
%
In the example \code{(b)}, the two assignments to \code{y} execute from a 
reaction to the same event \code{A}, and are simultaneous from an external 
point of view.
Since \CEU employs lexical order for intra-reaction statements, the execution 
is still deterministic, and \code{y} always becomes \code{(1+1)*2->4}.
%
However, any (apparently innocuous) change in the order of trails can change 
the semantics of a program, which we consider unsafe.

To mitigate this threat, \CEU performs a temporal analysis at compile time and 
detects concurrent accesses to shared variables, as follows:
if a variable is written in a trail segment, then a concurrent trail segment 
cannot read or write to that variable, nor dereference a pointer of that 
variable type.
%An analogous policy is applied for pointers \emph{vs} variables and pointers 
%\emph{vs} pointers.
%
A trail segment is a sequence of statements followed by an \code{await} (or 
termination).
Concurrency in \CEU is characterized when two or more trail segments in 
parallel react to the same input event.
%
Considering the examples in Figure~\ref{lst.shared}:
%
\begin{itemize}
\item The assignments to \code{x} in lines 2 and 5 in \code{(a)}
      \textbf{cannot} be concurrent because they are not in parallel trails.
\item The assignments to \code{x} in lines 5 and 8 in \code{(a)}
      \textbf{cannot} be concurrent because they \textbf{cannot} execute during 
      the same reaction.
\item The assignments to \code{y} in lines 5 and 8 in \code{(b)}
      \textbf{can} be concurrent because they are in parallel trails and 
      \textbf{can} execute during the same reaction.
\end{itemize}
%
The algorithm for the analysis, which is depicted in 
Section~\ref{sec.impl.analsyis}, inspects all possible \code{await} statements 
that precede a variable access and hold the associated awaking events.
Then, it checks all accesses in parallel trails to see if they share an awaking 
event.
If it is the case, the compiler warns about the suspicious accesses.

Note that this analysis is only possible due to the uniqueness of input events 
within reactions.
Otherwise, any two trail segments in parallel could be concurrent, even if they 
react to different input events.

\subsection{ Safe Integration with \C }
\label{sec.ceu.c}

In \CEU, any identifier prefixed with an underscore is repassed \emph{as is} to 
the \C compiler that generates the final binary.
Therefore, access to \C is seamless and, more importantly, easily trackable.
%
Similarly to Esterel, which supports the \code{call} primitive for external 
code, calls are assumed to be instantaneous~\cite{esterel.primer}.
%
Evidently, programs should only resort to \C for asynchronous functionality, 
such as non-blocking I/O, or simple \code{struct} accessors, but never for 
control purposes.

\begin{figure}[b]
\begin{minipage}[t]{0.50\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3.5em]
native do
    #define NUM 10
    void f  (void)  { <...> }
    void g  (int v) { <...> }
    int  id (int v) { <...> }
end
par/and do
    _f();
with
    _g(_id(_NUM));
end
\end{lstlisting}
\centering\small{(a)}
\end{minipage}
%
\begin{minipage}[t]{0.44\linewidth}
\begin{lstlisting}
native @const _NUM;
native @pure  _id;
native @safe  _f with _g;

.
\end{lstlisting}
\centering\small{(b)}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ TODO
\label{lst.c}
}
\end{figure}

In addition, \CEU also employs the temporal analysis for calls and accesses to 
external symbols in \C.
%
As an example, the program in Figure~\ref{lst.c}.a defines four external 
symbols inside a \code{native} block for standard declarations in \C (ln.  
1--6).
%
The \code{par/and} (ln. 7--11) creates two trails that react concurrently 
during the boot reaction (ln. 8,10):
the first trail calls the symbol \code{\_f},
while the second calls \code{\_g} and \code{\_id}, and also accesses 
\code{\_NUM}.
%
Since \CEU does not inspect any code defined in \C, it has no idea about the 
meaning of the symbols and complains about suspicious concurrent accesses 
between \code{\_f} against all symbols in the second trail.

The code in Figure~\ref{lst.c}.b uses annotations to provide hints to the 
compiler about the semantics of the \C symbols in use in \code{(a)}:
%
\begin{itemize}
\item \code{\_NUM} is a constant symbol, meaning that it is safe to use it 
      concurrently with any other symbol in the program.
\item \code{\_id} is a pure function, also meaning that it is safe to call it
      concurrently with any other symbol in the program.
\item Both \code{\_f} and \code{\_g} are impure, but their effects do not 
      conflict and they can be safely called concurrently.
\end{itemize}

Esterel's \code{abort} and \CEU's \code{par/or} statements illustrate how 
synchronous languages can abort awaiting lines of execution without tweaking 
them with synchronization primitives.
In contrast, traditional (asynchronous) multi-threaded languages cannot express 
thread termination safely~\cite{esterel.preemption,sync_async.threadsstop}.
%
Still, aborting lines of execution that deal with external resources is 
challenging because they may end up in an inconsistent state.
%
For this reason, Esterel and \CEU provide a \code{finalize} construct with a 
clause that executes even if the enclosing block aborts and does not terminate 
normally.

Figure~\ref{lst.fin} uses \code{lock} and \code{unlock} calls to represent an 
external resource.
The normal behavior is to \code{lock} the resource, perform some operations in 
subsequent reactions to input \code{A}, and then \code{unlock} the resource.
However, if the aborting input \code{B} occurs after the \code{lock} but before 
the reactions to \code{A}, we still want to call \code{unlock} to safely 
release the resource.
%
In Esterel and \CEU, the finalize clauses (ln. a:10 and b:5) are automatically 
called if the enclosing blocks (ln. a:3--1 and b:3--10) are externally aborted 
(ln. a:12 and b:12).

We completely missed the "finalize" construct of Esterel and recognize the 
similarity with our proposal.
We now emphasize that Céu tracks pointers entering and leaving the program to 
enforce appropriate use of "finalize" (as part of the section about interfacing 
with C).
